---
title: "Example of fitted values and orthogonal projections"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 200)
```

**This example will be due as part of your next homework assignment.  We're just getting a start on it in class together.**

### Model Statement

Suppose we use the model
\begin{align*}
y_i &= \beta + \varepsilon_i \\
\varepsilon_i &\sim \text{Normal}(0, \sigma^2)
\end{align*}

Also suppose we have $n = 2$ observations, and the observed response vector is $y = \begin{bmatrix} 1 \\ 2 \end{bmatrix}$.

#### (a) What is the design matrix $X$?

$X = \begin{bmatrix}1 \\ 1 \end{bmatrix}$

#### (b) Find the hat matrix $H$.

\begin{align*}
H &= X (X'X)^{-1} X' \\
&= \begin{bmatrix}1 \\ 1 \end{bmatrix} \left(\begin{bmatrix}1 & 1 \end{bmatrix} \begin{bmatrix}1 \\ 1 \end{bmatrix} \right)^{-1} \begin{bmatrix}1 & 1 \end{bmatrix} \\
&= \begin{bmatrix}1 \\ 1 \end{bmatrix}  \begin{bmatrix} 2 \end{bmatrix}^{-1} \begin{bmatrix}1 & 1 \end{bmatrix} \\
&= \begin{bmatrix}1 \\ 1 \end{bmatrix} \frac{1}{2} \begin{bmatrix}1 & 1 \end{bmatrix} \\
&= \begin{bmatrix}\frac{1}{2} & \frac{1}{2} \\ \frac{1}{2} & \frac{1}{2} \end{bmatrix} \\
\end{align*}

#### (c) Find the fitted values $\hat{y} = H y$.

\begin{align*}
\hat{y} &= H y \\
&= \begin{bmatrix} \frac{1}{2} & \frac{1}{2} \\ \frac{1}{2} & \frac{1}{2} \end{bmatrix} \begin{bmatrix} 1 \\ 2 \end{bmatrix} \\
&= \begin{bmatrix} 1.5 \\ 1.5 \end{bmatrix}
\end{align*}


#### (d) Draw a figure showing $\mathcal{C}(X)$ (it is a line), $y$, and $\hat{y}$, clearly labelling each.  Connect $y$ and $\hat{y}$ with a line segment, and by drawing an appropriate right angle on your figure, illustrate that $\hat{y}$ is the orthogonal projection of $y$ onto $\mathcal{C}(X)$.

The design matrix $X$ has one column, $\begin{bmatrix}1 \\ 1 \end{bmatrix}$.  $\mathcal{C}(X)$ is the line going through that point and the origin.  $y$ is the point $\begin{bmatrix} 1 \\ 2 \end{bmatrix}$, and as found in part (c), $\hat{y}$ is the point $\begin{bmatrix} `1.5 \\ 1.5 \end{bmatrix}$.  $\hat{y}$ is the orthogonal projection of $y$ into $\mathcal{C}(X)$, so the line connecting $\hat{y}$ and $y$ meets $\mathcal{C}(X)$ at a right angle.


```{r, echo = FALSE, fig.width=5, fig.height=4, warning=FALSE}
library(ggplot2)
ggplot() +
  geom_abline(intercept = 0, slope = 1) +
  geom_line(
    data = data.frame(x = c(1, 1.5), y = c(2, 1.5)),
    mapping = aes(x = x, y = y),
  ) +
  geom_line(
    data = data.frame(x = c(1.3, 1.4), y = c(1.5, 1.6)),
    mapping = aes(x = x, y = y),
  ) +
  geom_line(
    data = data.frame(x = c(1.3, 1.4), y = c(1.5, 1.4)),
    mapping = aes(x = x, y = y),
  ) +
  geom_point(
    data = data.frame(
      x = c(1, 1.5, 1),
      y = c(1, 1.5, 2),
      point = c("X", "y hat", "y")
    ),
    mapping = aes(x = x, y = y, color = point, shape = point),
    size = 2
  ) +
  scale_color_manual(values = c("black", "orange", "cornflowerblue")) +
  scale_x_continuous(limits = c(-3, 3)) +
  scale_y_continuous(limits = c(-3, 3)) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  xlab("Observation 1") +
  ylab("Observation 2") +
  theme_bw()
```